---
title: "Simulation Framework"
author: "Ross Gayler"
date: "2021-07-19"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

suppressPackageStartupMessages(
  {
    library(tibble)
    library(purrr)
    library(dplyr)
    library(ggplot2)
    library(tictoc)
  }
)

source(here::here("R", "functions.R"))
```

```{r, include = FALSE, cache = FALSE}
# Get all the functions defined in "R/functions.R"
# so we can display the source code in this notebook
# with a guarantee that they will be identical
# to the code executed here and in other notebooks

# See https://bookdown.org/yihui/rmarkdown-cookbook/read-chunk.html
# The argument to read_chunk() *must* be a literal string
knitr::read_chunk("R/functions.R")
```

This notebook documents the functions used to simulate the repeated
updating of the state of the PID controller.

The general issues around the software interface between the VSA and
multicopter components are discussed in [Relation of VSA and multicopter
simulation components](design_notes.html#sim-vsa-to-multicopter).

The specific software design issues around the simulation of the PID
controller are discussed in [Structure of the VSA data flow graph
implementation](design_notes.html#sim-dfg-design).

# `run_simulation()`

The function to run the simulation is defined as:

```{r, run_simulation, eval = FALSE}
```

The arguments are:

-   `input_df` - The history of the input variables.\
    This is a data frame with one row for each time step
    ($t = 1 \ldots n_{steps}$) and one column for each input variable.
    For this project the input variables are numeric scalars, but in
    general they can be arbitrarily complex structures if the column is
    a [list
    column](https://jennybc.github.io/purrr-tutorial/ls13_list-columns.html).
-   `init_state` - The initial state of the state variables.\
    This is a a one-row date frame. There is one column for each state
    variable. It *must* contain those state variables where the update
    function depends on the previous value of that variable. Those state
    variables which are VSA vectors *must* be list columns.
-   `f_update` - The state update function.\
    This is a function of two arguments. The first argument is a one-row
    data frame containing the previous values of the state variables. The
    second argument is a one-row data frame containing the input
    variables for that time step. The value of the state update function
    is a one-row data frame containing the updated values of the state
    variables.

The value of the `run_simulation()` function is a data frame with
columns for the input variables, state variables, and time ($t$). There
is one row for each of the time steps ($t = 1 \ldots n_{steps}$).

# Set up example

The following sections set up an example of using `run_simulation()`. This
is a drastically scaled down example in that the number of time steps
and VSA vector dimensions are greatly reduced so that the simulation
result can be viewed in its entirety.

## Set global parameters

Set the dimensionality of the VSA vectors and the number of time steps
in the simulation.

```{r}
v_dim <- 10 # vector dimension of state variables (in practice = 1e4)
n_step <- 5 # number of time steps in simulation (in practice = 500)
```

## Input variables

Create a data frame to hold the input values. Normally these will be
from a file exported from the multicopter simulation.

The input values are contained in a data frame.

Each of the input variables is a time series - one value per time step.
The data frame contains as many input variables as you need.

For this example we are just generating random input data. The input
data for this project are all numeric scalars, so it could have been
stored in a matrix, but in general, the inputs might be mixed types or
complex structures so we use a data frame for generality.

```{r}
input_df <- tibble::tibble(
  i_x = rnorm(n = n_step),
  i_y = rnorm(n = n_step)
)

input_df
```

## Initial state

The initial state consists of whatever state variables the state update
function refers to as previous state.

The previous state is represented as a one-row data frame with one
column for each of the required state variables. The VSA vector
variables need to be stored in [list
columns](https://jennybc.github.io/purrr-tutorial/ls13_list-columns.html).

I expect that we will generally use a function to construct the initial
state.

```{r}
# Function to make the initial state
mk_init_state <- function(
  v_dim, # integer[1] - dimension of vector-valued state variables
  seed = NULL # integer[1] - random generation seed
) # value - dataframe[1, ] - initial state of system
{
  set.seed(seed)
  
  # Construct initial state
  # MUST contain all columns used as previous state by update_state()
  # Any other state columns are optional
  tibble::tibble_row(
    # scalar valued state variables
    s_x = 0, s_y = 0,
    # vector valued state variables (each must be wrapped in list())
    s_0 = sample(c(-1, 1), size = v_dim, replace = TRUE) %>% list(), 
    s_1 = sample(c(-1, 1), size = v_dim, replace = TRUE) %>% list()
  )
}

# Make the initial state
init_state <- mk_init_state(v_dim, seed = 42)

init_state
str(init_state)
```

## State update function

The state update function takes the previous state and current input
variables as arguments and returns the current state.

This specific update function is a gibberish example, but is generically
the sort of thing I will be doing

The previous state is given as a one-row data frame. The VSA vector
values must be stored as [list
columns](https://jennybc.github.io/purrr-tutorial/ls13_list-columns.html).

The current input variables are given as a one-row data frame. For this
project the input variables are numeric scalars which are represented as
standard columns, but more generally the inputs may be more complex and
require list columns.

The value returned by the state update function is a one-row data frame
containing the updated values of all the state variables.

```{r}
# Function to update state from previous state and current input
update_state <- function(
  prev_state, # dataframe[1, ] - previous state
  input # dataframe[1, ] - current input
) 
{
  # Update scalar valued state variables
  s_x = prev_state$s_x + input$i_x 
  s_y = prev_state$s_y + input$i_y 
  
  # Update vector valued state variables
  # List columns must be unlisted before use
  s_0 = unlist(prev_state$s_0) * s_x
  s_1 = unlist(prev_state$s_1) * s_y
  s_2 = s_0 * s_1
  s_3 = s_2 * s_0
  s_4 = s_2 * s_1
  s_5 = s_3 + s_4
  
  # Construct and return a 1-row dataframe containing the updated state
  tibble::tibble_row(
    # Scalar valued state variables
    s_x, s_y, 
    # Vector valued state variables
    # List columns must be wrapped by list()
    s_0 = list(s_0),
    s_1 = list(s_1),
    s_2 = list(s_2),
    s_3 = list(s_3),
    s_4 = list(s_4),
    s_5 = list(s_5)
  )
}
```

# Run small simulation

Run the simulation based on the previous settings.

```{r}
# Run the simulation, returning the history as a data frame
history_df <- input_df %>% run_simulation(init_state, update_state)

# Take a look at the shape of the history
history_df
str(history_df)
```

# Run large simulation

Increase the parameters to realistic values and run the simulation.

```{r}
# Set global parameters
v_dim <- 1e4 # vector dimension of state variables
n_step <- 1e3 # number of time steps in simulation

# Generate input variables (random)
input_df <- tibble::tibble(
  i_x = rnorm(n = n_step),
  i_y = rnorm(n = n_step)
)

# Make the initial state
init_state <- mk_init_state(v_dim, seed = 42)

# Run the simulation, returning the history as a data frame
# Time the execution with tictoc::
tictoc::tic()
history_df <- input_df %>% run_simulation(init_state, update_state)
tictoc::toc()

# Take a look at the history
history_df

# How much RAM does the history use?
format(object.size(history_df), standard = "IEC", units = "GiB")
```

That's adequately fast.

The size is as expected.
