---
title: "VSA Basic Operators"
author: "Ross Gayler"
date: "2021-07-11"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

suppressPackageStartupMessages(
  {
    library(fs)
    library(here)
    library(magrittr)
  }
)

source(here::here("R", "functions.R"))
```

```{r, include = FALSE, cache = FALSE}
# Get all the functions defined in "R/functions.R"
# so we can display the source code in this notebook
# with a guarantee that they will be identical
# to the code executed here and in other notebooks

# See https://bookdown.org/yihui/rmarkdown-cookbook/read-chunk.html
# The argument to read_chunk() *must* be a literal string
knitr::read_chunk("R/functions.R")
```

This notebook ...

The reasoning behind the design choices is explained in
[XXX](design_notes.html#dfd-01).

# Generate atomic vectors

For this project we are using bipolar vectors ($V \in \{-1, +1\}^D$).

The vectors will be dense. That is, there will be no zero elements. We
will not be investigating the effect of sparsity in this project.

Define a function to create a randomly selected bipolar VSA vector.

```{r, vsa_mk_atom_bipolar, eval = FALSE}
```

The `seed` argument allows for a reproducible random selection.

The vector elements are integers rather than floats. This halves the
required storage space and is a minor gesture towards optimisation. Much
greater optimisation could be probably be achieved by using bit strings
rather than integers, but that's not worth the effort at this stage.

Where possible all other operations will be defined to accept floats
*and* integers to allow moving away from a strictly bipolar
representation if necessary.

Do some very small scale testing.

```{r}
v1 <- vsa_mk_atom_bipolar(10L)
v2 <- vsa_mk_atom_bipolar(10L)

v1
v2
all(v1 == v2)
```

-   Multiple calls generate different vectors.

```{r}
v1 <- vsa_mk_atom_bipolar(10L, seed = 1L)
v2 <- vsa_mk_atom_bipolar(10L, seed = 1L)

v1
v2
all(v1 == v2)
```

-   Setting the seed to the same value generates the same vector.

# Vector measures

We really only need the cosine similarity of two vectors. However,
define functions for the components of the cosine in case they are
useful.

These functions will work for any vectors. They do not have to be
bipolar vectors.

## Vector magnitude

Define a function to calculate the of a VSA vector.

```{r, vsa_mag, eval = FALSE}
```

I have not taken any numerical analysis considerations into account, so
don't hold any strong expectations for accuracy in extreme cases.

Do some very small scale testing.

```{r}
vsa_mag(0)
vsa_mag(1)
vsa_mag(2)
vsa_mag(-2)
```

-   The magnitude of a scalar is its absolute value.

```{r}
vsa_mk_atom_bipolar(9L) %>% vsa_mag()
vsa_mk_atom_bipolar(100L) %>% vsa_mag()
vsa_mk_atom_bipolar(1e4L) %>% vsa_mag()
vsa_mk_atom_bipolar(1e8L) %>% vsa_mag()
```

-   The magnitude of a bipolar vector is the square root of its
    dimensionality.

As the vector dimensionality is increased the operations take longer to
execute and eventually something will break, e.g. you will run out of
RAM or the dimensionality will be too large to be represented as an
`integer`.

```{r}
(vsa_mk_atom_bipolar(100L) * 1L) %>% vsa_mag()
(vsa_mk_atom_bipolar(100L) * 1.3) %>% vsa_mag()
(vsa_mk_atom_bipolar(100L) * -5) %>% vsa_mag()
```

-   Rescaling the vector is equivalent to rescaling its magnitude
    ($\Vert kV \Vert = \vert k \vert \Vert V \Vert$).

## Vector dot product

Define a function to calculate the dot product of two VSA vectors.

```{r, vsa_dotprod, eval = FALSE}
```

I have not taken any numerical analysis considerations into account, so
don't hold any strong expectations for accuracy in extreme cases.

Do some very small scale testing.

```{r}
vsa_dotprod(1, 1)
vsa_dotprod(1, 0)
vsa_dotprod(1, 3)
vsa_dotprod(2, 3)
vsa_dotprod(-2, 3)
vsa_dotprod(-2, -3)
```

-   The dot product of two scalars is their product.

```{r}
vsa_dotprod(vsa_mk_atom_bipolar(9L, seed = 42L), vsa_mk_atom_bipolar(9L, seed = 42L))
vsa_dotprod(vsa_mk_atom_bipolar(100L, seed = 43L), vsa_mk_atom_bipolar(100L, seed = 43L))
vsa_dotprod(vsa_mk_atom_bipolar(1e4L, seed = 44L), vsa_mk_atom_bipolar(1e4L, seed = 44L))
vsa_dotprod(vsa_mk_atom_bipolar(1e8L, seed = 45L), vsa_mk_atom_bipolar(1e8L, seed = 45L))
```

-   The dot product of a vector with itself is equal to the square of
    its magnitude.
-   The dot product of a bipolar vector with itself is equal to its
    dimensionality.

```{r}
vsa_dotprod(vsa_mk_atom_bipolar(1e4L), vsa_mk_atom_bipolar(1e4L))
vsa_dotprod(vsa_mk_atom_bipolar(1e4L), vsa_mk_atom_bipolar(1e4L))
vsa_dotprod(vsa_mk_atom_bipolar(1e4L), vsa_mk_atom_bipolar(1e4L))
vsa_dotprod(vsa_mk_atom_bipolar(1e4L), vsa_mk_atom_bipolar(1e4L))
```

-   The dot product of two randomly selected high dimensional vectors is
    approximately zero (relative to the dimensionality).

## Vector cosine similarity

Define a function to calculate the cosine of the angle between two VSA
vectors.

```{r, vsa_cos_sim, eval = FALSE}
```

I have not taken any numerical analysis considerations into account, so
don't hold any strong expectations for accuracy in extreme cases.

Do some very small scale testing.

Don't bother looking at pairs of scalars, because although the cosine of
the angle between them is defined it is not interesting. (The vectors
are parallel by definition.)

```{r}
vsa_cos_sim(vsa_mk_atom_bipolar(9L, seed = 42L), vsa_mk_atom_bipolar(9L, seed = 42L))
vsa_cos_sim(vsa_mk_atom_bipolar(100L, seed = 43L), vsa_mk_atom_bipolar(100L, seed = 43L))
vsa_cos_sim(vsa_mk_atom_bipolar(1e4L, seed = 44L), vsa_mk_atom_bipolar(1e4L, seed = 44L))
vsa_cos_sim(vsa_mk_atom_bipolar(1e8L, seed = 45L), vsa_mk_atom_bipolar(1e8L, seed = 45L))
```

-   The cosine of the angle of a vector with itself is one. (The vectors
    are coincident.)

```{r}
vsa_cos_sim(vsa_mk_atom_bipolar(1e4L, seed = 42L), 1 * vsa_mk_atom_bipolar(1e4L, seed = 42L))
vsa_cos_sim(vsa_mk_atom_bipolar(1e4L, seed = 43L), 2 * vsa_mk_atom_bipolar(1e4L, seed = 43L))
vsa_cos_sim(vsa_mk_atom_bipolar(1e4L, seed = 44L), -2 * vsa_mk_atom_bipolar(1e4L, seed = 44L))
```

-   Scaling has no effect on the direction of a vector
-   The sign of scaling does effect the sign of the direction of a
    vector.

```{r}
vsa_cos_sim(vsa_mk_atom_bipolar(1e4L), vsa_mk_atom_bipolar(1e4L))
vsa_cos_sim(vsa_mk_atom_bipolar(1e4L), vsa_mk_atom_bipolar(1e4L))
vsa_cos_sim(vsa_mk_atom_bipolar(1e4L), vsa_mk_atom_bipolar(1e4L))
vsa_cos_sim(vsa_mk_atom_bipolar(1e4L), vsa_mk_atom_bipolar(1e4L))
```

-   The dot product of two randomly selected high dimensional vectors is
    approximately zero.

# Negate vector

Define a function to calculate the negation of a VSA vector. This simply
reverses the direction of the vector.

```{r, vsa_negate, eval = FALSE}
```

This function will work for any vector. It does not have to be a bipolar
vector.

Do some very small scale testing.

```{r}
# make some example vectors
v1 <- vsa_mk_atom_bipolar(1e4L)
v2 <- vsa_mk_atom_bipolar(1e4L)

vsa_mag(v1) == (vsa_negate(v1) %>% vsa_mag())
vsa_mag(v2) == (vsa_negate(v2) %>% vsa_mag())
```

-   Negation does not change the magnitude of a vector.

```{r}
vsa_cos_sim(v1, vsa_negate(v1))
vsa_cos_sim(v2, vsa_negate(v2))

vsa_cos_sim(v1, v2)
vsa_cos_sim(v1, vsa_negate(v2))
```

-   Negation reverses the direction of a vector.

# Permute vector

Define function to permute a vector.

These will be functions to:

-   Randomly generate the specification of a permutation
-   Generate the specification for the inverse of a given permutation
-   Apply a permutation to a VSA vector

## Generate permutation

Define a function to randomly generate the specification of a
permutation.

```{r, vsa_mk_perm, eval = FALSE}
```

The `seed` argument allows for a reproducible random selection.

The returned value is a vector of integers of length `vsa_dim`. This is
a random permutation of the integers `1 : vsa_dim`. When applied as a
permutation each element of the specification indicates the input
element to be copied to that position in the output.

Do some very small scale testing.

```{r}
p1 <- vsa_mk_perm(10L, seed = 1L)
p2 <- vsa_mk_perm(10L, seed = 1L)

p1
p2
all(p1 == p2)
```

-   Setting the seed to the same value generates the same permutation.

```{r}
p1 <- vsa_mk_perm(10L)
p2 <- vsa_mk_perm(10L)

p1
p2
all(p1 == p2)
```

-   Multiple calls generate different permutations

## Generate inverse permutation

Define a function to generate the inverse of a given permutation.

```{r, vsa_mk_inv_perm, eval = FALSE}
```

The returned value is a vector of integers of length `length(perm)`.
This is a permutation of the integers `1 : length(perm)`. When applied
as a permutation each element of the specification indicates the input
element to be copied to that position in the output.

Do some very small scale testing.

```{r}
# make some example permutation specifications
p_identity <- 1L:10L
p_reverse <- seq(10L, 1L, -1L)
p_left1 <- c(p_identity[-1], p_identity[1])
p_right1 <- c(p_identity[10], p_identity[-10])
p_random <-vsa_mk_perm(10L)

p_identity
vsa_mk_inv_perm(p_identity)
```

-   The identity permutation is its own inverse.

```{r}
p_reverse
vsa_mk_inv_perm(p_reverse)
vsa_mk_inv_perm(p_reverse) %>% vsa_mk_inv_perm()
```

-   The reverse permutation is its own inverse.

```{r}
p_left1
vsa_mk_inv_perm(p_left1)
p_right1
vsa_mk_inv_perm(p_left1) %>% vsa_mk_inv_perm()
```

-   The inverse of left cyclic shift is right cyclic shift.
-   The inverse of the inverse transform is the identity transform

```{r}
p_random
vsa_mk_inv_perm(p_random)
vsa_mk_inv_perm(p_random) %>% vsa_mk_inv_perm()
```

-   The inverse of the inverse transform is the identity transform

## Apply permutation

Define a function to apply the given permutation to a VSA vector.

```{r, vsa_permute, eval = FALSE}
```

The returned value is the input VSA vector permuted according to the
specified permutation.

Do some very small scale testing.

```{r}
# make some example VSA vectors
v1 <- 1L:10L
v2 <- 101L:110L

v1
v1 %>% vsa_permute(p_identity)
v2
v2 %>% vsa_permute(p_identity)
```

-   The identity permutation leaves the VSA vectors unchanged.

```{r}
v1
v1 %>% vsa_permute(p_reverse)
v2
v2 %>% vsa_permute(p_reverse)
```

-   The reverse permutation reverses the VSA vectors.

```{r}
v1
v1 %>% vsa_permute(p_left1)
v2
v2 %>% vsa_permute(p_left1)
```

-   The left cyclic shift permutation performs a left cyclic shift of
    the VSA vectors.

```{r}
v1
v1 %>% vsa_permute(vsa_mk_inv_perm(p_left1))
v2
v2 %>% vsa_permute(vsa_mk_inv_perm(p_left1))
```

-   The inverse of the left cyclic shift permutation performs a right
    cyclic shift of the VSA vectors.

```{r}
v1
v1 %>% vsa_permute(p_random)
v1 %>% vsa_permute(p_random) %>% vsa_permute(vsa_mk_inv_perm(p_random))
```

-   The the composition of a random permutation and its inverse are
    equivalent to the identity permutation.

# Multiply vectors

Define a function to multiply an arbitrary number of VSA vectors.

```{r, vsa_multiply, eval = FALSE}
```

Do some very small scale testing.

```{r}
v1 <- 1:5
v2 <- 2:6

v1
v2
vsa_multiply(v1, v1)
vsa_multiply(v1, v1, v2)
```

-   Multiplication is the element-wise product of the argument vectors.

```{r}
vsa_multiply(1, 1)
vsa_multiply(1, -1)
vsa_multiply(-1, -1)
```

-   Signs work as expected.

# Add vectors

Vector addition in this project is performed by randomly sampling
elements from each of the source VSA vectors. The sampling specification
(the VSA vector argument to use as the source for each element of the
output VSA vector) can be randomised per applicatioon of the addition
function or can be set to a constant across applications of the addition
function.

## Generate sampling specification

Define a function to generate a sampling specification for addition of
an arbitrary number of VSA vectors.

The VSA vector dimensionality and the number of vectors to be added must
be specified.

The sampling weights correspond to the VSA vector arguments to the
addition. The sampling weights must be non-negative and are normalised to
sum to one so that the can be treated as probabilities.

```{r, vsa_mk_sample_spec, eval = FALSE}
```

Do some very small scale testing.

```{r}
vsa_mk_sample_spec(vsa_dim = 10L, sample_wt = c(1, 1, 1), seed = 42L)
vsa_mk_sample_spec(vsa_dim = 10L, sample_wt = c(1, 1, 1), seed = 42L)
```

-   Setting the seed makes the sampling sepcification reproducible.

```{r}
vsa_mk_sample_spec(vsa_dim = 10L, sample_wt = c(1, 0))
vsa_mk_sample_spec(vsa_dim = 10L, sample_wt = c(0, 1, 1))
vsa_mk_sample_spec(vsa_dim = 10L, sample_wt = c(1, 3))
```

-   The sample weights are converted to probabilities of selection.

## Apply sampling specification

Define a function to calculate the weighted sum of an arbitrary number
of VSA vectors.

```{r, vsa_add, eval = FALSE}
```

Do some very small scale testing.

```{r}
# create vectors to add
# make unique values so they can be uniquely tracked
x1 <- 10L:19L
x2 <- 20L:29L
x3 <- 30L:39L

# specify the sampling
vsa_add(x1,x2,x3, sample_spec = c(1L,2L,3L,1L,2L,3L,1L,2L,3L,1L))
vsa_add(x1,x2,x3, sample_spec = c(1L,2L,3L,1L,2L,3L,1L,2L,3L,1L))
```

-   Sampling is fixed when `sample_spec` is specified.

```{r}
vsa_add(x1,x2,x3, sample_wt = c(0, 0, 1))
```

-   Extreme random weighting works as expected.

```{r}
vsa_add(x1,x2,x3)
vsa_add(x1,x2,x3)
vsa_add(x1,x2,x3)
```

-   Randomised sampling is different on every occasion.
